Sentence Generation as Planning with Probabilistic LTAG\\

### 2010-05-19 trunk/Papers/pcrisp-10/crisp.tex
Many local changes, restructured section 4

\subsection{Sentence Generation in CRISP}
Like {\sc spud}, {\sc crisp} uses an {\sc ltag} in which elementary trees are assigned semantic content. Each node in a {\sc crisp} elementary tree node is associated with a semantic role. Semantic content is expressed as a set of literals, encoding relations between these roles. All nodes that dominate the lexical anchor are assigned the role `self', which intuitively corresponds to the event or individual described by this tree. Figure \ref{fig:grammar}(a) shows an example grammar of this type.

In a derivation we may only include elementary trees whose semantic content has an instantiation in the knowledge base. For each substitution and adjunction, the semantic role associated with the role of the target node is unified with the `self' role of the child tree. For example, given the knowledge base 
{\it \{cat($i_2$), fish($i_3$), raw($i_3$), eats($i_1$,$i_2$,$i_3$)\}} and the grammar in \ref{fig:grammar}(a), {\sc crisp} could produce the derivation in \ref{fig:grammar}(b).

\subsection{CRISP Planning Domains} 
\label{ssec:crispdomain}
Before we describe {\sc crisp}'s encoding of sentence generation as planning, we briefly review AI planning in general. 
The crucial concept in AI planning is a state, a conjunction of first order literals describing relations between some individuals. A planning problem consist of an initial state, a set of goal states and a set of planning operators that describe possible state transitions.  A solution to the problem is any sequence of actions (instantiated operators) that leads from the initial state to one of the goal states. Planning problems can be written in a standard notation ({\sc pddl}\footnote{Planning Domain Description Language}, \newcite{gerevinietal2009}) and can be solved efficiently by planning systems such as {\sc ff}\footnote{Fast Forward} \cite{hoffmannnebel2001}.

In {\sc crisp} planning states correspond to partial {\sc tag} derivations and record open substitution and adjunction sites, semantic individuals associated with them, and parts of the communicative goal that have not yet been expressed.  The initial state also encodes the knowledge base and the communicative goal. Each planning operator contributes a new elementary tree to the derivation and at the same time satisfies part of the communicative goal, as described in the previous section. In a goal state there are no open substitution sites left and all literals in the communicative goal have been expressed. 
\begin{figure}[t]
\begin{center}
\planaction{\bf S-eats-1(u,~x1,~x2,~x3)}{step(step1),referent(u,~x1),\\ subst(S,~u), eats(x1,~x2,~x3)}{
$\lnot$needtoexpr(pred-eats,~x1,~x2,~x3),\\ $\lnot$subst(S,~u),\\
subst(NP,~subj-1), subst(NP,~obj-1),\\ refer(subj-1,~x2), refer(obj-1,~x3)\\
canadj(VP, u), canadj(V, u), canadj(S, u)\\
$\lnot$ step(step1), step(step2)}\\\smallskip

\planaction{\bf NP-cat-2(u,~x1)}{step(step1),referent(u,~x1),\\ subst(NP,~u), cat(x1)}{
$\lnot$needtoexpress(pred-cat,~x1),\\ $\lnot$subst(NP,~u),\\
canadj(N, u), canadj(NP, u)\\
$\lnot$ step(step1), step(step2)}\\\smallskip

\planaction{\bf N-raw-4(u,~x1)}{step(step1),referent(u,~x1),\\ canadj(N,~u), raw(x1)}{
$\lnot$needtoexpress(pred-raw,~x1),\\ $\lnot$canadj(N,~u),\\
$\lnot$ step(step4), step(step5)}\\\smallskip
\end{center}
\caption{\label{fig:crisp-operators} {\sc crisp} operators for some of the elementary trees in figure \ref{fig:grammar}.}
\end{figure}

@@

\subsection{CRISP and Large Grammars}
Trees in large, possibly treebank induced grammars are often applicable in a number of different syntactic contexts, leading to overgeneration. Such trees contain many internal nodes that are open for adjunction but will not necessarily be filled in a derivation for a given input communicative goal. Attaching to the wrong node can produce invalid or disprefered word order. 
In addition, in such grammars the ambiguity of possible, structurally different trees (supertags) for a given word is high. These problems are illustrated in figure \ref{fig:overgen}. Assuming a grammar that includes trees for both right attaching ($t13$) and left attaching PPs ($t252$), both derivations (a) and (b) are {\it grammatical} derivations that satisfy the same communicative goal. However, most readers disprefer the reading in (b). 
Clearly, to use {\sc crisp} with such a grammar we need a method of distinguishing good derivations from bad ones. 
\begin{figure*}
\begin{center}
\includegraphics[width=.6\textwidth]{figures/overgen.pdf}
\caption{\label{fig:overgen} Two derivations with a large grammar, that satisfy the same communicative goal. Sentence (b) is disprefered by most readers.}
\end{center}
\end{figure*}

>>>>

\subsection{Sentence Generation in CRISP}
Like {\sc spud}, {\sc crisp} uses an {\sc ltag} in which elementary trees are assigned semantic content. Each node in a {\sc crisp} elementary tree is associated with a semantic role. Semantic content is expressed as a set of literals, encoding relations between these roles. All nodes that dominate the lexical anchor are assigned the role `self', which intuitively corresponds to the event or individual described by this tree. Figure \ref{fig:grammar}(a) shows an example grammar of this type.

In a derivation we may only include elementary trees whose semantic content has an instantiation in the knowledge base. For each substitution and adjunction, the semantic role associated with the role of the target node is unified with the `self' role of the child tree. For example, given the knowledge base 
{\it \{cat($i_2$), fish($i_3$), raw($i_3$), fish($i_4$), eats($i_1$,$i_2$,$i_3$)\}} and the grammar in \ref{fig:grammar}(a), {\sc crisp} could produce the derivation in \ref{fig:grammar}(b). Notice that {\sc crisp} can generate the unambiguos refering expression `the raw fish' for $i_3$, to distinguish it from $i_4$. For details on  refering expression generation in {\sc crisp} see \newcite{garoufikoller2010}.

\subsection{CRISP Planning Domains} 
\label{ssec:crispdomain}
Before we describe {\sc crisp}'s encoding of sentence generation as planning, we briefly review AI planning in general. 
A planning state is a conjunction of first order literals describing relations between some individuals. A planning problem consist of an initial state, a set of goal states and a set of planning operators that describe possible state transitions.  A solution to the problem is any sequence of actions (instantiated operators) that leads from the initial state to one of the goal states. Planning problems can be solved efficiently by general purpose planning systems such as 
{\sc ff} \cite{hoffmannnebel2001}.  
% Planning problems can be written in a standard notation ({\sc pddl}\footnote{Planning Domain Description Language}, \newcite{gerevinietal2009}).

In {\sc crisp} planning states correspond to partial {\sc tag} derivations and record open substitution and adjunction sites, semantic individuals associated with them, and parts of the communicative goal that have not yet been expressed.  The initial state also encodes the knowledge base and the communicative goal. Each planning operator contributes a new elementary tree to the derivation and at the same time satisfies part of the communicative goal, as described in the previous section. In a goal state there are no open substitution sites left and all literals in the communicative goal have been expressed. 
\begin{figure}[t]
\begin{center}
\planaction{\bf subst-t28-eats-S(u,~x1,~x2,~x3)}{referent(u,~x1),\\ subst(S,~u), eats(x1,~x2,~x3)}{
$\lnot$needtoexpr(pred-eats,~x1,~x2,~x3),\\ $\lnot$subst(S,~u),\\
subst(NP,~subj-1), subst(NP,~obj-1),\\ referent(subj-1,~x2), referent(obj-1,~x3)\\
canadj(VP, u), canadj(V, u), canadj(S, u)}\\\smallskip

\planaction{\bf subst-t3-cat-NP(u,~x1)}{referent(u,~x1),\\ subst(NP,~u), cat(x1)}{
$\lnot$needtoexpr(pred-cat,~x1),\\ $\lnot$subst(NP,~u),\\
canadj(N, u), canadj(NP, u)}\\ \smallskip

\planaction{\bf subst-t5-raw-N(u,~x1)}{referent(u,~x1),\\ canadj(N,~u), raw(x1)}{
$\lnot$needtoexpr(pred-raw,~x1),\\ $\lnot$canadj(N,~u)}\\\smallskip
\end{center}
\caption{\label{fig:crisp-operators} {\sc crisp} operators for some of the elementary trees in figure \ref{fig:grammar}.}
\end{figure}

@@

\subsection{CRISP and Large Grammars}
Using {\sc crisp} with large, treebank induced grammars can lead to overgeneration, 
%Such trees contain many internal nodes that are open for adjunction but will not necessarily be filled in a derivation for a given input communicative goal. Attaching to the wrong node can produce invalid or disprefered word order. 
due to the high lexical ambiguity of possible, structurally different trees (supertags) for a given word. We illustrate this problem in figure \ref{fig:overgen}. Assuming a grammar that includes trees for both right adjoining ($t13$) and left adjoining PPs ($t252$), both derivations (a) and (b) are {\it grammatical} derivations that satisfy the same communicative goal. However, most readers disprefer the reading in (b). 
Clearly, to use {\sc crisp} with such a grammar we need a method of distinguishing good derivations from bad ones. In section \ref{sec:experiments} we provide empirical evidence for this problem and show how it is solved by the approach we describe in the next section. 
\begin{figure}
\begin{center}
\includegraphics[width=.3\textwidth]{figures/overgen.pdf}
\caption{\label{fig:overgen} Two derivations with a large grammar, that satisfy the same communicative goal. Sentence (b) is disprefered by most readers.}
\end{center}
\end{figure}

### 2010-05-19 trunk/Papers/pcrisp-10/experiments.tex
Many local changes, restructured section 4

\section{Experiments}
\label{sec:experiments}
In this section we perform experiments with the statistical generation system described in the previous section.  We are interested in the output quality and runtime behavior of different combinations of planning systems and probability models. We can assume that the sentence planning abilities are inherited from {\sc crisp} and only evaluate the realization abilities of our system by comparing generated sentences against a gold standard testset. The next section describes our evaluation data and section \ref{ssec:eval-measures} describes the evaluation measures used.

@@

For our experiments we use a grammar and treebank, that was automatically extracted from the Wall Street Journal portion of the Penn Treebank (41576 sentences)
using the algorithm described by \newcite{chenschanker2004}.  We preprocess this grammar by linearizing multiple adjunctions and by re-attaching prepositions to trees they appear with.

>>>>

\section{Experiments}
\label{sec:experiments}
In this section we perform experiments with the statistical generation system described in the previous section.  We are interested in the output quality and runtime behavior of different combinations of planning systems and probability models. Like {\sc crisp}, {\sc pcrisp} is a sentence generation system and can for instance generate refering expressions. Unfortunately, since we do not have an evaluation dataset for full sentence generation, we only evaluate the realization abilities of our system by comparing generated sentences against a gold standard testset. The next section describes our evaluation data and section \ref{ssec:eval-measures} describes the evaluation measures used.

@@

For our experiments we use a grammar and treebank that was automatically extracted from the Wall Street Journal using the algorithm described by \newcite{chenschanker2004}.

This algorithm outputs a grammar that allows multiple adjunctions at the same node. For such a grammar {\sc ptag} is not a suitable probability model. Models that can deal with multiple adjunction are discussed by \newcite{nasrrambow2006}, but employing them would require non-trivial modifications to our encoding as a metric planning problem. We therefore preprocess the treebank by `linearizing' multiple adjunctions. Furthermore, we attach prepositions to the trees they substitute in, to increase the expressiveness of the bilexical probability distribution. For details on these preprocessing steps see \cite{bauer2009}.

### 2010-05-19 trunk/Papers/pcrisp-10/pcrisp.tex
Many local changes, restructured section 4

We now extend {\sc crisp} to statistical generation ({\sc pcrisp}). The basic idea is to add a statistical grammar model while leaving the sentence generation mechanism untouched. This way we can select the highest scoring derivation, that satisfies all constraints (grammaticality, expresses the communicative goal, use unambiguous refering expressions etc.). We first propose three probability models in the next section and adapt {\sc crisp} planning problems to acomodate probabilities in section \ref{ssec:pcrisp-domains}.

\subsection{Probability Models}
\label{ssec:probmodels}

{\bf Probabilistic TAG:}
As a basic probability model over {\sc ltag} derivations we use probabilistic {\sc tag}\footnote{we refer to this formalism as \sc{pltag}} \cite{resnik1992}, 
which provides a probability for each substitution and adjunction event in a derivation. In addition, it assigns a probability to the event of starting a derivation with a specific initial tree and to the event of not adjoining anything to an open adjunction node at all. Assuming that the probabilities of these events within a derivation are independent of each other, the probability of a derivation is then simply defined as the product of all elementary event probabilities.
Formally,

@@

Our choice of {\sc pltag} for sentence generation is motivated by a number of attractive properties.
 First, as {\sc pltag} is lexicalized, it does not only assign probabilities to operations in the grammar (as for example plain {\sc pcfg}), but also accounts for binary dependencies between words. Unlike n-gram models however, these co-occurences are structured according to local syntactic context as a result of {\sc tag}'s extended domain of locality. The probability model describes how the syntactic arguments of a word are typically filled. 
Furthermore, as {\sc tag} factors recursion from the domain of dependencies, the probability for core constructions remains the same independent of additional adjunctions. 
Second, {\sc pltag} is a generative model. This is crucial because it allows us to assign probabilities to partial derivations. In addition we use the probability of partial derivations as a cost function, to help guide search through the space of possible PLTAG derivations. In contrast, other statistical generation systems use discriminative models that select the best possible generation output from a set of candidate sentences.
Finally, the independence assumption makes it easy to estimate {\sc pltag}s from a treebank.

To alleviate data sparseness, caused by the specific event definition of {\sc pltag}, we propose two alternative probability models. The first model is an unlexicalized version of {\sc pltag}, the second model uses a simple smoothing approach.\\ 
{\bf Unlexicalized Probabilities:}
  An easy way to deal with data sparseness is to drop all lexicalization from event definitions, as illustrated in figure \ref{fig:modelillustration}, A. Unfortunately the model does no longer account for bilexical dependencies between words. Since our system has to add a lexicalized tree in each step, lexicalization for this child tree should always be taken into account by the probability model, if available. Despite these drawbacks, we perform experiments with the unlexicalized model as a baseline. This allows us to investigate if purely syntactic information is sufficient to achieve high quality generation output. \\  
{\bf Linear Interpolation:}
 The third model computes linear interpolation between three back-off levels. The first level is just standard {\sc pltag} (figure \ref{fig:modelillustration}, B.1), for the second level the lexicalization of the parent tree is dropped (figure \ref{fig:modelillustration}, B.2), for the third level the model describes only the distribution of lexicalized child trees over each category (figure \ref{fig:modelillustration}, B.3). 
The third level is similar in spirit to a probabilistic version of the original generation as planning formulation in \cite{kollerstone2007}.
\begin{figure}[t]
\begin{center}
\includegraphics[width=.6\textwidth]{figures/modelillustration}
\caption{\label{fig:modelillustration} Illustration of the unlexicalized probability model (A) and the three back-off levels of the linear interpolation model (B). B.1 corresponds to the original {\sc pltag} definition.}

@@

\subsection{PCRISP Planning Domains}
\label{ssec:pcrisp-domains}
In this section we reformulate the {\sc crisp} planning operators described in section \ref{ssec:crispdomain}.
The independence assumption in {\sc pltag} allows us to maintain that planning operator add a single tree, but are now assigned a probability score. However, while {\sc crisp} planning operators can add an elementary tree to any site of the right category, {\sc pltag} events are binary events between lexicalized trees at a specific node. We therefore adapt the literals that record open substitution and adjunction sites in partial derivations accordingly and create one operator for each node in each possible combination of lexicalized trees.

Different approaches to planning with numeric values exist. We can encode probabilities either as timespans required by each operator and use a temporal planner, or as as generic numeric variable that is increased by each operator. In both cases this value (which can be seen as a cost) is summed over actions and a resulting plan should seek to minimize it. We therefore set the cost of an operator to be its negative log probability. 
Figure \ref{example-action} shows an example planning operator. We adopt this scheme for all planning operators regardless of the probability model.
\begin{figure}[t]
\begin{center}
\cplanaction{\bf subst-t3-cat-t28-eats-n2(u,~x1)}{step(step1),referent(u,~x1),\\ subst(t28-eats,n2,~u), cat(x1)}{
$\lnot$needtoexpress(pred-cat,~x1),\\ $\lnot$subst(t-28-eats, n2,~u),\\
adj(t3-cat, n0, u), adj(t3-cat, n1 u)\\
$\lnot$ step(step1), step(step2)}{4.3012}\\\smallskip
\caption{\label{example-action} {\sc pcrisp} Operator to substitute $t3$ lexicalized with 'cat' for node 2 of $t28$ lexicalized with `eats'.}
\end{center}
\end{figure}
%\newcommand{\init}[0]{\textit{ init}}
%\newcommand{\subst}[0]{\textit{ subst}}

>>>>

We now extend {\sc crisp} to statistical generation ({\sc pcrisp}). The basic idea is to add a statistical grammar model while leaving the sentence generation mechanism untouched. This way we can select the highest scoring derivation, that satisfies all constraints (grammaticality, expresses the communicative goal, use unambiguous refering expressions etc.).

 As a straightforward probability model over {\sc ltag} derivations we choose probabilistic {\sc tag} ({\sc ptag}) \cite{resnik1992}.
Our choice of {\sc ptag} for sentence generation is motivated by a number of attractive properties.
 {\sc Ptag} is lexicalized and therefore does not only assign probabilities to operations in the grammar (as for example plain {\sc pcfg}), but also accounts for binary dependencies between words.  Unlike n-gram models however, these co-occurences are structured according to local syntactic context as a result of {\sc tag}'s extended domain of locality. The probability model describes how the syntactic arguments of a word are typically filled. 
Furthermore, as {\sc tag} factors recursion from the domain of dependencies, the probability for core constructions remains the same independent of additional adjunctions. 
%Second, {\sc ptag} is a generative model. This is crucial because it allows us to assign probabilities to partial derivations. In addition we use the probability of partial derivations as a cost function, to help guide search through the space of possible PLTAG derivations. In contrast, other statistical generation systems use discriminative models that select the best possible generation output from a set of candidate sentences.
%Finally, the independence assumption makes it easy to estimate {\sc ptag}s from a treebank.
 We review {\sc ptag} in section \ref{ssec:probmodels}.

While we leave the basic sentence generation mechanism intact, we need to modify the concrete formulation of {\sc crisp} planning operators, as each of them now corresponds to some probability event. Likewise, we need to take the step from classical planning to metric planning systems, that can use the probabilities.
In metric planning \cite{fox2002}, instead of changing only boolean state variables, planning actions can decrease or increase a numeric variable. The goal state specifies constraints on this variable. In the simplest case the variable can only be increased by a static cost value in each action, and the goal state contains the objective to minimize the total cost. Systems such as metric-{\sc FF} \cite{hoffmann2003} do not guarantee optimality, but generally offer good results. We address our encoding of sentence generation with {\sc ptag} as metric planning in section \ref{ssec:pcrisp-domains}.

\subsection{Probabilistic TAG}
\label{ssec:probmodels}
{\sc ptag} \cite{resnik1992} views {\sc tag} derivations as sequences of probability events, each corresponding to a substitution or adjunction. In addition, it assigns a probability to the event of starting a derivation with a specific initial tree and to the event of not adjoining anything to an open adjunction node at all. Assuming that the probabilities of these events within a derivation are independent of each other, the probability of a derivation is then simply defined as the product of all elementary event probabilities. 
Formally,

@@

\subsection{PCRISP Planning Domains}
\label{ssec:pcrisp-domains}
Here, we reformulate the {\sc crisp} planning operators described in section \ref{ssec:crispdomain}.
The independence assumption in {\sc ptag} allows us to maintain that planning operator add a single tree, but are now assigned a probability score. However, while {\sc crisp} planning operators can add an elementary tree to any site of the right category, {\sc ptag} events are binary events between lexicalized trees at a specific node. We therefore adapt the literals that record open substitution and adjunction sites in partial derivations accordingly and create one operator for each node in each possible combination of lexicalized trees.
Figure \ref{example-action} shows an example planning operator. We adopt this scheme for all planning operators regardless of the probability model.
\begin{figure}[t]
\begin{center}
\cplanaction{\bf subst-t3-cat-t28-eats-n2(u,~x1)}{referent(u,~x1),\\ subst(t28-eats,n2,~u), cat(x1)}{
$\lnot$needtoexpress(pred-cat,~x1),\\ $\lnot$subst(t-28-eats, n2,~u),\\
adj(t3-cat, n0, u), adj(t3-cat, n1 u)}{4.3012}\\\smallskip
\caption{\label{example-action} {\sc pcrisp} Operator to substitute $t3$ lexicalized with 'cat' for node 2 of $t28$ lexicalized with `eats'.}
\end{center}
\end{figure}

Finally, we set the cost of an operator to be its negative log probability. For example for the operator in figure \ref{example-action} $$Cost( \subst(t_3,\textit{`cat'},t_{28},\textit{`eats"},n_2) ) =$$
$$ -\log( P(\subst(t_3,\textit{`cat'},t_{28},\textit{`eats"},n_2)) ). $$ This way the plan with the smallest cost sum corresponds to the {\sc tag} derivation with the highest probability.

\subsection{Dealing with Data Sparseness}
\label{ssec:sparseness}
%To alleviate data sparseness, caused by the specific event definition of {\sc ptag}, we propose two alternative probability models. The first model is an unlexicalized version of {\sc ptag}, the second model uses a simple smoothing approach.\\ 
The probability event definition of {\sc ptag} is very fine grained. Substitution and adjunction events depend on specific parent and child trees with specific lexicalizations and on a node in the parent tree, as illustrated in figure \ref{fig:modelillustration}, B.1. When we estimate a probability model from training data, we cannot expect to observe evidence for all combinations of trees. Derivations that include such unseen events, have zero probability and are therefore impossible. As we show in section \ref{sec:experiments}, this gives rise to a massive data sparseness problem.

  A straightforward way to deal with data sparseness is to drop all lexicalization from event definitions, as illustrated in figure \ref{fig:modelillustration}, A. Unfortunately this model does no longer account for bilexical dependencies between words. Since our system has to add a lexicalized tree in each step, lexicalization for this child tree should always be taken into account by the probability model, if available. Despite these drawbacks, we perform experiments with the unlexicalized model as a baseline. This allows us to investigate if purely syntactic information is sufficient to achieve high quality generation output. \\  
An alternative model computes linear interpolation between three back-off levels. The first level is just standard {\sc ptag} (figure \ref{fig:modelillustration}, B.1), for the second level the lexicalization of the parent tree is dropped (figure \ref{fig:modelillustration}, B.2), for the third level the model describes only the distribution of lexicalized child trees over each category (figure \ref{fig:modelillustration}, B.3).  
Notice also, that the third level is similar in spirit to a probabilistic version of original {\sc crisp} operators (compare figure \ref{fig:modelillustration}, B.3 and figure \ref{fig:crisp-operators}. 
\begin{figure}[t]
\begin{center}
\includegraphics[width=.5\textwidth]{figures/modelillustration}
\caption{\label{fig:modelillustration} Illustration of the unlexicalized probability model (A) and the three back-off levels of the linear interpolation model (B). B.1 is the original {\sc ptag} definition and corresponds directly to {\sc pcrisp} operators as in figure \ref{example-action}, while B.3 resembles {\sc crisp} operators (c.f. figure \ref{fig:crisp-operators}).}

@@

%\newcommand{\init}[0]{\textit{ init}}
%\newcommand{\subst}[0]{\textit{ subst}}
